---
title: "The Max Step"
---


# The GEV Model


$$
\mathrm{GEV(y|\mu, \sigma, \xi)} = \begin{cases}
e^{- \left(1 + \xi \frac{y - \mu}{\sigma}\right)_+^{-1/\xi}}, \xi\neq0 \\
e^{- e^{- \frac{y - \mu}{\sigma}
}}, \xi=0
\end{cases}
$$

In our case we want $\mu$ to vary with time. If we write $y_{it}$ for the observed hourly maximum at station $i$ during year $t$ we will then have

$$
y_{it} \sim \mathrm{GEV}(\mu_{it}, \sigma_i, \xi_i), \quad \mu_{it}, \sigma_i > 0, \quad \xi_i \in (-0.5, 0.5),
$$

where

$$
\mu_{it} = \mu_i \cdot (1 + \Delta_i \cdot (t - t_0)),
$$

and

$$
1 + \xi_i \frac{y_{it} - \mu_{it}}{\sigma_i} > 0, \quad \forall i, t.
$$

# Maximum Likelihood Estimation

## Transformed parameters

Having performed the Max step and saved the ML estimates we can easily load them by fetching the object `station_estimates`. Here we plot the distribution of transformed estimates.

$$
\begin{aligned}
\psi &= \log(\mu) \\
\tau &= \log(\frac{\sigma}{\mu}) = \log(\sigma) - \log(\mu) \\
\phi &= \mathrm{logit}(\xi, \text{ scale}=1.6) - 0.8 \\
\gamma &= \mathrm{logit}(\Delta, \text{ scale}=0.02) - 0.01.
\end{aligned}
$$

These transformations make it so that $\xi$ can take values between $-0.8$ and $0.8$, and $\Delta$ can take values between $-0.01$ and $0.01$. 

# Results

```{r}
#| message: false
#| warning: false
library(bggjphd)
library(tidyverse)
library(GGally)
library(cowplot)
library(arrow)
library(here)
theme_set(theme_bggj())
```

```{r}
d <- station_estimates |> 
  select(station, par) |> 
  unnest(par) |> 
  inner_join(
    stations
  ) |> 
  select(station, proj_x, proj_y, name, value)
```

## Distributions

### Transformed Scale

```{r}
d |> 
  ggplot(aes(value)) +
  geom_histogram(bins = 60) +
  facet_wrap("name", scales = "free") +
  labs(
    x = NULL,
    y = NULL,
    title = "Distributions of GEV parameters from Max step"
  )
```

```{r}
#| fig-asp: 1
d |> 
  pivot_wider() |> 
  select(-station, -proj_x, -proj_y) |> 
  ggpairs(progress = FALSE)
```


### Original Scale

```{r}
d |> 
  pivot_wider() |> 
  mutate(mu = exp(psi),
         sigma = exp(tau + psi),
         xi = link_shape_inverse(phi),
         delta = link_trend_inverse(gamma)) |> 
  select(-psi, -tau, -phi, -gamma, -proj_x, -proj_y) |> 
  pivot_longer(c(-station)) |> 
  mutate(name = fct_relevel(name, "mu", "sigma", "xi", "delta")) |> 
  ggplot(aes(value)) +
  geom_histogram(bins = 100) +
  facet_wrap("name", scales = "free") +
  labs(
    x = NULL,
    y = NULL,
    title = "Distributions of backtransformed GEV parameters from Max step"
  )
```


```{r}
#| fig-asp: 1
d |> 
  pivot_wider() |> 
  mutate(mu = exp(psi),
         sigma = exp(tau + psi),
         xi = link_shape_inverse(phi),
         delta = link_trend_inverse(gamma)) |> 
  select(-psi, -tau, -phi, -gamma, -station, -proj_x, -proj_y) |> 
  ggpairs(progress = FALSE)
```


## Spatial Distribution

### Location

```{r}
d |> 
  filter(name == "psi") |> 
  ggplot(aes(proj_x, proj_y, fill = value)) +
  geom_raster(interpolate = TRUE) +
  scale_x_continuous(
    expand = expansion(),
    breaks = c(range(d$proj_x), pretty(d$proj_x))
  ) +
  scale_y_continuous(
    expand = expansion(),
    breaks = c(range(d$proj_y), pretty(d$proj_y))
  ) +
  scale_fill_viridis_c() +
  labs(
    x = "X Projection",
    y = "Y Projection",
    fill = NULL,
    title = "Spatial distribution of Psi"
  )
```

```{r}
d |> 
  filter(name == "psi") |> 
  ggplot(aes(proj_x, proj_y, fill = exp(value))) +
  geom_raster(interpolate = TRUE) +
  scale_x_continuous(
    expand = expansion(),
    breaks = c(range(d$proj_x), pretty(d$proj_x))
  ) +
  scale_y_continuous(
    expand = expansion(),
    breaks = c(range(d$proj_y), pretty(d$proj_y))
  ) +
  scale_fill_viridis_c() +
  labs(
    x = "X Projection",
    y = "Y Projection",
    fill = NULL,
    title = "Spatial distribution of Mu"
  )
```

### Scale

```{r}
d |> 
  filter(name == "tau") |> 
  ggplot(aes(proj_x, proj_y, fill = value)) +
  geom_raster(interpolate = TRUE) +
  scale_x_continuous(
    expand = expansion(),
    breaks = c(range(d$proj_x), pretty(d$proj_x))
  ) +
  scale_y_continuous(
    expand = expansion(),
    breaks = c(range(d$proj_y), pretty(d$proj_y))
  ) +
  scale_fill_viridis_c() +
  labs(
    x = "X Projection",
    y = "Y Projection",
    fill = NULL,
    title = "Spatial distribution of Tau"
  )
```

```{r}
d |> 
  pivot_wider() |> 
  mutate(sigma = exp(tau + psi)) |> 
  ggplot(aes(proj_x, proj_y, fill = sigma)) +
  geom_raster(interpolate = TRUE) +
  scale_x_continuous(
    expand = expansion(),
    breaks = c(range(d$proj_x), pretty(d$proj_x))
  ) +
  scale_y_continuous(
    expand = expansion(),
    breaks = c(range(d$proj_y), pretty(d$proj_y))
  ) +
  scale_fill_viridis_c() +
  labs(
    x = "X Projection",
    y = "Y Projection",
    fill = NULL,
    title = "Spatial distribution of Sigma"
  )
```

### Shape

```{r}
d |> 
  filter(name == "phi") |> 
  ggplot(aes(proj_x, proj_y, fill = value)) +
  geom_raster(interpolate = TRUE) +
  scale_x_continuous(
    expand = expansion(),
    breaks = c(range(d$proj_x), pretty(d$proj_x))
  ) +
  scale_y_continuous(
    expand = expansion(),
    breaks = c(range(d$proj_y), pretty(d$proj_y))
  ) +
  scale_fill_viridis_c() +
  labs(
    x = "X Projection",
    y = "Y Projection",
    fill = NULL,
    title = "Spatial distribution of Phi"
  )
```

```{r}
d |> 
  pivot_wider() |> 
  mutate(xi = link_shape_inverse(phi)) |> 
  ggplot(aes(proj_x, proj_y, fill = xi)) +
  geom_raster(interpolate = TRUE) +
  scale_x_continuous(
    expand = expansion(),
    breaks = c(range(d$proj_x), pretty(d$proj_x))
  ) +
  scale_y_continuous(
    expand = expansion(),
    breaks = c(range(d$proj_y), pretty(d$proj_y))
  ) +
  scale_fill_viridis_c() +
  labs(
    x = "X Projection",
    y = "Y Projection",
    fill = NULL,
    title = "Spatial distribution of Xi"
  )
```

### Trend

```{r}
d |> 
  filter(name == "gamma") |> 
  ggplot(aes(proj_x, proj_y, fill = value)) +
  geom_raster(interpolate = TRUE) +
  scale_x_continuous(
    expand = expansion(),
    breaks = c(range(d$proj_x), pretty(d$proj_x))
  ) +
  scale_y_continuous(
    expand = expansion(),
    breaks = c(range(d$proj_y), pretty(d$proj_y))
  ) +
  scale_fill_viridis_c() +
  labs(
    x = "X Projection",
    y = "Y Projection",
    fill = NULL,
    title = "Spatial distribution of Gamma"
  )
```

```{r}
d |> 
  pivot_wider() |> 
  mutate(delta = link_trend_inverse(gamma)) |> 
  ggplot(aes(proj_x, proj_y, fill = delta)) +
  geom_raster(interpolate = TRUE) +
  scale_x_continuous(
    expand = expansion(),
    breaks = c(range(d$proj_x), pretty(d$proj_x))
  ) +
  scale_y_continuous(
    expand = expansion(),
    breaks = c(range(d$proj_y), pretty(d$proj_y))
  ) +
  scale_fill_viridis_c() +
  labs(
    x = "X Projection",
    y = "Y Projection",
    fill = NULL,
    title = "Spatial distribution of Delta"
  )
```
